---
title: "Practical Machine Learning Assignment"
author: "Yoshiteru Kageyama"
date: "November 14, 2015"
output: html_document
---
## Executive summary
In this assignment, I downloaded the personal activity data that were derived
from the source http://groupware.les.inf.puc-rio.br/har. I applied the data
to a machine learning algorithm, random forest model to predict how a participant
exercised with a dumbbell, which was classified in five categories(class A to E). 
This algorithm resulted in the very precise prediction to the five categories.

## Read the training and testing data
First, I read the given training and testing data from the Internet.

```{r warning=FALSE, cache=TRUE}
training_data <- 'pml-training.csv'
testing_data <- 'pml-testing.csv'
if (!file.exists(training_data)) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                destfile=training_data)
}
if (!file.exists(testing_data)) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                destfile=testing_data)
}
pml_train <- read.csv(file=training_data, header=T)
pml_test  <- read.csv(file=testing_data, header=T)
```

## Cleaning the training data

I found that several columns in the pml_train data frame were regarded as a factor.
After looking into them, I found that they contained blanks and some strange values,
for example, "!DIV/0".
So, I converted them into numeric values forcefully.

```{r warning = FALSE, cache=TRUE}
for(col in c("kurtosis_roll_belt", "kurtosis_picth_belt", "kurtosis_yaw_belt",
             "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
             "max_yaw_belt", "min_yaw_belt",
             "kurtosis_roll_arm", "kurtosis_picth_arm", "kurtosis_yaw_arm",
             "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
             "kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell",
             "skewness_roll_dumbbell", "skewness_pitch_dumbbell", "skewness_yaw_dumbbell",
             "max_yaw_dumbbell", "min_yaw_dumbbell",
             "amplitude_yaw_belt","amplitude_yaw_dumbbell",
             "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
             "skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm",
             "max_yaw_forearm", "min_yaw_forearm",
             "amplitude_yaw_forearm"
             )) {
  colNum <- levels(pml_train[,col])[pml_train[,col]]
  pml_train[,col] <- as.numeric(colNum)
  colNum <- levels(pml_test[,col])[pml_test[,col]]
  pml_test[,col] <- as.numeric(colNum)
}
#summary(pml_train)
```

When I tried to apply these data to some machine learning algorithms, using the
columns that contained a lot of NAs wasn't appropriate. So I omitted them in advance.

```{r cache=TRUE}
pml_train <- pml_train[, colSums(!is.na(pml_train)) == nrow(pml_train)]
pml_test <-pml_test[, colSums(!is.na(pml_test)) == nrow(pml_test)]
```

Furthermore, the following columns should be omitted: 

- user_name
- timestamp values
- new_window and num_window
- total acceleration values and their standard deviation

Because these variables don't have anything to do with predicting the classe in
applying a machine learning algorithm.

```{r cache=TRUE}
pml_train <- pml_train[, -which(colnames(pml_train) %in% c("X", "user_name",
                            "raw_timestamp_part_1", "raw_timestamp_part_2",
                            "cvtd_timestamp", "new_window", "num_window",
                            "total_accel_belt", "total_accel_arm", "total_accel_dumbbell"
                            ))]
pml_test <- pml_test[, -which(colnames(pml_test) %in% c("X", "user_name",
                            "raw_timestamp_part_1", "raw_timestamp_part_2",
                            "cvtd_timestamp", "new_window", "num_window",
                            "total_accel_belt", "total_accel_arm", "total_accel_dumbbell"
                            ))]
summary(pml_train)
```

This summary showed us that all data were appropriate to apply some machine
learning algorithm (there isn't any NA, and all the data were well-balanced).

The meaning of the classe column is as follows:

| Class    |   Meaning                               |
|:--------:|:----------------------------------------|
|   A      | Exactly according to the specification  |
|   B      | Throwing the elbows to the front        |
|   C      | Lifting the dumbbell only halfway       |
|   D      | Lowering the dumbbell only halfway      |
|   E      | Throwing the hips to the front          |

## Select and create the machine learning model

At first, I selected the random forest as a learning model for the training data,
because generally speaking, the randowm forest algorithm is suitable to fit
the classification problem like this assignment.
I tried to create data partition to pml_train data and fit the new training data
to the random forest model.(It took a lot of time...)

```{r cache=TRUE}
pml_train.inTrain <- createDataPartition(y=pml_train$classe, p=0.7, list=FALSE)
pml_train.train   <- pml_train[pml_train.inTrain,]
pml_train.test    <- pml_train[-pml_train.inTrain,]
modFit <- train(classe ~ ., data=pml_train.train, 
                method="rf", ntree=100)
```

Now I checked if the created model is accurate enough to predict the classe using
the test cases take from the training data.

```{r}
pml_train.predict <- predict(modFit, pml_train.test)
table(pml_train.predict, pml_train.test$classe)
pml_train.predict.accuracy <- 
  sum(pml_train.predict == pml_train.test$classe) / length(pml_train.predict)
```

The accuracy rate of the model I've just created is as follows:
```{r}
print(pml_train.predict.accuracy)
```

I believe this result is encouraging enough to use it as a predictor for the
test data.

# Predicting The Results

So, I applied the acquired learning model to the 20 test data.

```{r}
# this function was taken from the Coursera instruction page.
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

# try to predict answers
answers <- as.character(predict(modFit, newdata=pml_test))
pml_write_files(answers)
```
I submitted the prediction results into the "Prediction Assignment Submission" page
on the Coursera and found that all of them were correct.